\jaxcosmo: An End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-11-10 paper_OJA.tex
colorJEC
some modifs following FL

%%% color Comments 
%% JEC : \FL produces a clash we another definition (but coannot figure out which one)  
\newcommand{\FrL}[1]{{\color{cyan}FL: #1}}
\newcommand{\JZ}[1]{{\color{purple}JZ: #1}}
\newcommand{\JEC}[1]{{\color{green}JEC: #1}}

@@

The second approach is to compute numerical derivatives by finite differences. This approach can be applied on black-box models of abitrary complexity (from Boltzmann codes \FrL{some citations} to cosmological simulations \FrL{some citations}). However it is notoriously difficult to obtain stable derivatives by finite differences \FrL{cite papers trying to do fisher}. In addition, their computational cost does not scale well with the number of parameters (a minimum of $(2N+1)$ model evaluations is typically required for $N$ parameters), making them completely inpractical whenever derivatives are needed on the fly.\FrL{this last sentence is not great.}

@@

Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class. \FrL{class inheritance is TMI.}  After the forward model creation for the DES-Y1 3x2pts exercise described in Appendix~\ref{app-DESY1} \FrL{We should introduce the model at the beginning of this section.}, we have performed 
\begin{itemize}
    \item first a reparametrisation of the latent variable uniform priors (Table~\ref{tab-DESY1}) to use a unified uniform prior $\mathcal{U}[-5,5]$; \FrL{these are not latent variables, but model parameters. Also people don't know what this reparametrisation means.}
    \item second for the warm-up phase \FrL{People don't know what the warmup phase is...} we ask for mass matrix optimisation with essentially a diagonal form but with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$. This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_cobaya_NUTS_SVI_bis}.
\end{itemize}
The NUTS sampler have been running with 16 chains of 1,000 samples each (\texttt{vectorized} option) \FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.} after a warm-up phase consisting of 200 samples, and the \texttt{max\_tree\_depth} has been set to 7 by default (ie. 128 steps for each iteration). The experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}. The elapse time for these experiments was 20 hours.

@@

\FrL{we don't have plots of just the VI?}

@@

%  - Massively Parallel MCMC on GPU (linked to vmap and/or pmap)
%  - Efficiency of gradient-based inference methods
Another aspect worth discussing are the prospects for scaling up and speeding up cosmological inference in practice given tools such as \jaxcosmo\ . As we illustrated in the previous section, gradient-based inference techniques yield significantly less correllated MCMC chains.\FrL{Hummm here I guess we want to say also it scales better with dimensions, and that you can get very fast first results.} \FrL{we also want to say that the GPU parallelisation is going to change the game, all the more when you account for new massive GPU machines.}

>>>>

%%% color Comments 
%% JEC : \FL produces a clash we another definition (but coannot figure out which one)  
\newcommand{\FrL}[1]{{\color{cyan}FL: #1}}
\newcommand{\JZ}[1]{{\color{purple}JZ: #1}}
\newcommand{\JEC}[1]{{\color{magenta}JEC: #1}}

@@

The second approach is to compute numerical derivatives by finite differences. This approach can be applied on black-box models of abitrary complexity (from Boltzmann codes \FrL{some citations} to cosmological simulations \FrL{some citations}). However it is notoriously difficult to obtain stable derivatives by finite differences \FrL{cite papers trying to do fisher}. In addition, their computational cost does not scale well with the number of parameters (a minimum of $(2N+1)$ model evaluations is typically required for $N$ parameters), making them completely impractical whenever derivatives are needed on the fly.\FrL{this last sentence is not great.}

@@

Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class. \FrL{class inheritance is TMI.}  After the forward model creation for the DES-Y1 3x2pts exercise described in Appendix~\ref{app-DESY1} \FrL{We should introduce the model at the beginning of this section.},
\begin{itemize}
    \item \st{a reparametrisation of the latent variable uniform priors to use a unified uniform prior $\mathcal{U}[-5,5]$} \JEC{we have first applyied  an affine transformation on cosmological, intrinsic alignment and biais parameters to use an unique uniform prior $\mathcal{U}[-5,5]$ (Table~\ref{tab-DESY1})}; \FrL{these are not latent variables, but model parameters. Also people don't know what this reparametrisation means.}
    \item  \FrL{People don't know what the warmup phase is...}\JEC{we use a structured mass matrix consisting on essentially a diagonal form  with a block structure for the following sets of parameters} $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$. This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_cobaya_NUTS_SVI_bis}.
\end{itemize}
The NUTS sampler have been running with 16 chains of 1,000 samples each (\texttt{vectorized} option) \FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.} after a warm-up phase consisting of 200 samples, and the \texttt{max\_tree\_depth} has been set to 7 by default (ie. 128 steps for each iteration). The experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}. The elapse time for these experiments was 20 hours.

@@

\FrL{we don't have plots of just the VI?}\JEC{SVI alone (ie. not followed by HMC) gives poor results.}

@@

%  - Massively Parallel MCMC on GPU (linked to vmap and/or pmap)
%  - Efficiency of gradient-based inference methods
Another aspect worth discussing are the prospects for scaling up and speeding up cosmological inference in practice given tools such as \jaxcosmo\ . As we illustrated in the previous section, gradient-based inference techniques yield significantly less correlated MCMC chains.\FrL{Hummm here I guess we want to say also it scales better with dimensions, and that you can get very fast first results.} \FrL{we also want to say that the GPU parallelisation is going to change the game, all the more when you account for new massive GPU machines.}
